---
title: "class11 Structural Bioinformatics"
output: github_document
---

## The PDB database

Let's examine the PDB structure database composition
Loaded PDB data summary using more show folder in new window and dragged into workspace.
```{r}
#Loading PDB summary into workspace and making it look nice. (Making first column row names)
stats <- read.csv("PDB Data Export Summary (1).csv", row.names = 1)
stats
```

Q. What percentage of strucutres solved by Cryo-EM

```{r}
pre.by.method <- stats$Total/sum(stats$Total) * 100
names(pre.by.method) <- row.names(stats)
pre.by.method
```

Q. What proportion of PDB structures are protein

```{r}
protein_percentage_pdb <- round(sum(stats$Proteins)/sum(stats$Total) *100, 1)
protein_percentage_pdb
```

How to get data without CSV file? Copy data and use Addins --> Paste as dataframe function from datapasta.

```{r}
data.frame(stringsAsFactors=FALSE,
   Experimental.Method = c("X-Ray", "Other", "NMR", "Multi Method",
                           "Electron Microscopy", "Total"),
              Proteins = c(124770, 250, 10988, 127, 2057, 138192),
         Nucleic.Acids = c(1993, 4, 1273, 5, 31, 3306),
    Protein.RNA.Complex = c(6451, 6, 257, 2, 723, 7439),
                 Other = c(10, 13, 8, 1, 0, 32),
                 Total = c(133224, 273, 12526, 135, 2811, 148969)
)

```

##Section 3 Bio3D

```{r}
#check if package is loaded by looking through packages tab.
library(bio3d)

```

Load HIV protease + inhibitor pdb file
```{r}
pdb <- read.pdb("1hsg")
pdb
```

Load bio3D-view package & view pdb file in R studio
```{r}
library("bio3d.view")

view(pdb, "overview", col = "sse")
```

Extract only protein only portion of this structure and write it out to a new pdb file.

Extract Ligand (i.e. HIV protease inhibitor)
```{r}
Inds <- atom.select(pdb, "ligand")
Inds
#Ligand.only <- trim.pdb(protein.atoms)
#write.pdb(protein.only)
```

```{r}
#Inds$atom
pdb$atom[Inds$atom, ]
```


```{r}
ligand.pdb <- trim.pdb(pdb, Inds)
ligand.pdb
```

```{r}
view(ligand.pdb)
```

```{r}
write.pdb(ligand.pdb, file = "ligand.pdb")
```


Lets make a protein only file now. function is 
read.pdb("1HSG")
Inds <- atom.select(pdb, "protein")
protein.pdb <- trim.pdb(PDB, Inds)
write.pdb(protein.pdb, file = "protein.pdb")


```{r}
Inds.protein <- atom.select(pdb, "protein")
Inds.protein
```

```{r}
protein.pdb <- trim.pdb(pdb, Inds.protein)
protein.pdb
```

```{r}
write.pdb(protein.pdb, file = "protein.pdb")
```


```{r}
pdb <- read.pdb("1HEL")
pdb

#Normal Mode Analysis
modes <- nma(pdb)
plot(modes)
```

```{r}
m7 <- mktrj(modes,
 mode=7,
 file="mode_7.pdb")
view(m7, col=vec2color( rmsf(m7) ))

```

