---
title: "Class 9 PCA and Clustering Project"
author: "Matt Maxwell"
date: "February 8, 2019"
output: html_document
---

## Machine Learning Cancer Biopsy Analysis

Put dataset file in my class09 directory then clicked file and selected "import dataset". This gave me a nice formatted dataset.


Could also use read.csv to read data set. Need to learn to do this!
```{r}
fna.data <- "WisconsinCancer (2).csv"
read.csv(fna.data)
wisc.df <- read.csv(fna.data)
```



Always double check your data after loading.
```{r}
head(wisc.df)
```


Looks like there is a funny last col "X". Lets check how many samples (i.e patients) .

```{r}
# number of patients
nrow(wisc.df)
```


```{r}
ncol(wisc.df)
```

Let's take rows 3 to 32 for further analysis 
```{r}
wisc.data <- wisc.df[ ,3:32]
head(wisc.data)
View(wisc.data)
```

Set row names to patient ID number.
```{r}
row.names(wisc.data) <- wisc.df$id
head(wisc.data)
```

Q. How many cancer/non-cancer samples?
```{r}
table(wisc.df$diagnosis)
```


Q. How many features/columns are "_mean" values?
```{r}
#colnames(wisc.data)
inds <- grep("_mean", colnames(wisc.data))
colnames(wisc.data)[inds]
```



Create a diagnosis vector.
```{r}
diagnosis <- as.numeric(wisc.data)
```



Calculate mean of every column. apply(x, margin, function). 
round(x, significant figures)
```{r}
round( apply(wisc.data, 2, mean), 2)
```

calculate stdv of every column.
```{r}
round( apply(wisc.data, 2, sd), 2)
```



## Principal Component Analysis
prcomp() is our function for principal component analysis.
prcomp(wisc.data, scale = TRUE)

```{r}
#Perform PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

Plot PCA analysis
```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[, 2], col = wisc.df$diagnosis, xlab = "wisc.pr$x[, 1]", ylab = "wisc.pr$x[, 2]")
```

Calculate variance of each component. Shows us why we chose our PC1 & PC2.

```{r}
pr.var <- wisc.pr$sdev^2
pve <- round( pr.var/sum(pr.var) * 100, 2)
```

```{r}
plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained",ylim = c(0, 1), type ="o")
```

Bar plot of PC.
```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```



##Hierarchical clustering

For hclust we need a distance matrix and we get this from our PCA results (i.e. wisc.pr$x)

```{r}
d <- dist(wisc.pr$x[, 1:2])
hc <- hclust(d, method = "ward.D2")
plot(hc)

```

Let's cut tree into 3 groups
```{r}
grp3 <- cutree(hc, k=3)
table(grp3)
plot(grp3)
```

Plot our PCA plot colored by cluster

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], xlab = "PC1", ylab = "PC2", col = grp3)
```


cross tabulation?

```{r}

#diagnosis 
diagnosis <- wisc.df$diagnosis == "M"

table(grp3, diagnosis)

```

Let's predict cancer status of new data
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grp3)
points(npc[,1], npc[,2], col= c("green", "blue"), pch=15, cex=3)
```

