---
title: "Class 14 Transcriptomics and Analysis of RNA-seq Data"
output: github_document
---

Study profiling ASM cell lines treated with dexamethasone focused in on gene CRISPLD2

#RNA seq Analysis
#Import countData and colData

```{r}
counts <- read.csv("Data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("Data/airway_metadata.csv", stringsAsFactors = FALSE)

```

Check the data
```{r}
head(counts)
```

```{r}
head(metadata)
```

```{r}
View(metadata)

```

```{r}
nrow(counts)
ncol(counts)

```

Check for metadata to counts correspondence
```{r}
all( metadata$id == colnames(counts)[-1] )
```
Our header ensgene is being counted as a column. use [-1] following colnames(counts) to match up our columns.


#Control counts
#this labels control samples counts with Ensemble gene identifiers
```{r}
control.ids <- metadata$id[metadata[,"dex"]=="control"]
control <- counts[,control.ids]
rownames(control) <- counts$ensgene
head(control)
```

#Treated counts
```{r}
treated.ids <- metadata$id[metadata[,"dex"] =="treated"]
treated <- counts[,treated.ids]
rownames(treated) <- counts$ensgene
head(treated)


```
Lets get the means for control and treated groups for all our genes.
```{r}
control.mean <- rowMeans(control)
head(control.mean)
```


```{r}
treated.mean <- rowMeans(treated)
head(treated.mean)
```


Get mean count value for each gene for both groups. Use data.frame() functuion for making this pretty

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```


```{r}
colSums(meancounts)
```

```{r}
plot(meancounts, col=c("red"))
```

#Use log scale for x & y axis!
```{r}
plot(meancounts, col=c("red", "blue"), log="xy")
```
#Create a Log2FC column in meancounts to examine fold change of treated samples relative to control.

```{r}
meancounts$log2fc <- log2(meancounts$treated.mean/meancounts$control.mean)
head(meancounts)

```




The negative value means they are below the 45˚ line if we look at the graph. It means when the cell is treated, the gene has a lower value than in control. -Inf means it's trying to log O. We want to filter all the 0 values.

Also, let's call our corrected meancounts "mycounts"
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```
arr.ind returns results as an array
```{r}
toy <- c(1,2,0,4,0)
which(toy==0)

newtoy <- cbind(toy,toy)
which(newtoy==0, arr.ind = TRUE)

unique( which(newtoy==0, arr.ind = TRUE)[,1])
```

#mycounts has corrected Logfold change column.
Now that we have cleaned up the data let's set up some threshold to help us furthur characterize the darta. It's common to set the threshold of greater than 2 or less than -2.
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

250 upregulated genes.
```{r}
sum(up.ind)
```
367 downregulated genes
```{r}
sum(down.ind)
```

##Let's add anotation data from Ensemble

```{r}
anno <- read.csv("Data/annotables_grch38.csv")
head(anno)
```

```{r}
dim(anno)
```

##Let's try-out the **merge()** function to add the annotation data to our **mycounts** (i.e., gene names to the ensemble identifiers)
```{r}
head(mycounts)
```

```{r}
mycounts.anno <- merge(x=mycounts, y=anno, by.x="row.names", by.y="ensgene")
head(mycounts.anno)
```

```{r}
source("http://bioconductor.org/biocLite.R")
biocLite("AnnotationDbi")
```

```{r}
source("http://bioconductor.org/biocLite.R")
biocLite("org.Hs.eg.db")
library(org.Hs.eg.db)
```

```{r}
columns(org.Hs.eg.db)

```

#Add columns for gene symbol, entrez id, and uniprot ID.
```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
mycounts$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

```

```{r}
mycounts$uniprot <- mapIds(org.Hs.eg.db,
                                    keys=row.names(mycounts),
                                    column="UNIPROT",
                                    keytype="ENSEMBL",
                                    multiVals="first")
```

```{r}
head(mycounts)
```

##DESeq2 Analysi
```{r}
library(DESeq2)
```
DESeq2 uses a special object called DESeqDataSet which is composed from the count data, the metadata and the experimental design. It will also store the data in the object.

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design= ~dex,
                              tidy= T)
```

```{r}
dds
```

Run DESeq Analysis
```{r}
dds <- DESeq(dds)
```


Get results from DESeq Analysi

```{r}
res <- results(dds)
res
```


```{r}
summary(res)
```
Let's change the p value from .1 to 0.05, more stringent cutoff.
```{r}
summary(res, alpha=0.05)
```

##Volcano plots
```{r}
plot(res$log2FoldChange, -log(res$padj))
```

Let's colorize the graph

#ran into trouble following entrez additon to mycounts. 
```{r}
mycols <- rep("grey", nrow(res))
mycols[abs(res$log2FoldChange) > 2] = "red"
subsuet.inds <- res$padj<0.01 & abs(res$log2FoldChange)>2

mycols[subsuet.inds] = "blue"
plot(res$log2FoldChange, -log(res$padj), col=mycols)
```

Let's do this in ggplot too, make some pretty graphs in ggplot.

```{r}
library(ggplot2)
ggplot(as.data.frame(res), aes(log2FoldChange, -log10(pvalue), col=mycols)) + 
    geom_point() + 
    ggtitle("Volcano plot")
```

Weird. Bioconductor package not available for the latest version of R.
```{r}
biocLite("EnhancedVolcano")

```
```{r}
library(EnhancedVolcano)
```

##Let's see about some box plots for CRISPLD2